% Course-provided IEEE Overleaf template structure:
% IEEEtran conference class, original two-column layout, and no manual
% font-size or line-spacing changes.
\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\graphicspath{{figures/}{report/figures/}}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{url}
\usepackage{hyperref}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Natural Disaster Severity Prediction via\\
Leakage-Aware Boosted Trees and Temporal Feature Engineering\\
{\footnotesize Data Mining Final Project, 2026 Spring}
}

\author{\IEEEauthorblockN{Shin-Yu Chen}
\and
\IEEEauthorblockN{Velda Wei-Hsin Hung}
\and
\IEEEauthorblockN{Sky Shih-Kai Hong}
}

\maketitle

\begin{abstract}
Group ID: 5. Members: Shin-Yu Chen (314551141,
sychen.cs14@nycu.edu.tw), Velda Wei-Hsin Hung (314553021,
velda.cs14@nycu.edu.tw), and Sky Shih-Kai Hong (314554024,
sky.cs14@nycu.edu.tw). Our code is available at GitHub:
\href{https://github.com/skyhong2002/disaster-severity-prediction}{\color{blue}https://github.com/skyhong2002/disaster-severity-prediction}.

This project predicts weekly drought severity scores for 2,248 regions from
historical meteorological observations. The central difficulty is not only
multi-horizon forecasting, but also preventing target leakage from weekly score
labels while adapting to train-test season and weather shift. We implement
direct horizon models using LightGBM, XGBoost, and CatBoost, with calendar,
lag, rolling, exponentially weighted, drought-proxy, climatology, and
91-day-gapped score-history features. The best reproducible reportable
submission is a 35/35/30 LightGBM/XGBoost/CatBoost ensemble with Kaggle public
MAE \textbf{0.8124}. A separate already-submitted public-chase horizon hedge
reaches public MAE \textbf{0.7905} and crosses Baseline 3, but we keep it
separate from the method claim because it was selected by public leaderboard
feedback rather than by a new training algorithm.
\end{abstract}

\section{Project Summary (\textcolor{red}{max 1 page})}

\subsection{Goal}
The competition task is to predict the drought severity score for each region
over the next five consecutive weeks after the 91-day test observation window.
Scores are on a 0--5 scale, with larger values indicating more severe drought.
The submission must contain one row per \texttt{region\_id} and five numeric
columns, \texttt{pred\_week1} through \texttt{pred\_week5}. The evaluation
metric is mean absolute error (MAE).

\subsection{Approach}
We frame the task as supervised direct multi-step regression. For each horizon
$h \in \{1,\ldots,5\}$, the training label is the real weekly score $h$ weeks
after a forecast origin. The final pipeline builds a time-safe feature table
from raw weather, recent and long-window weather summaries, calendar variables,
region-month climatology anomalies, and score-history features shifted at
least 91 days. The final reportable model is a convex ensemble of three
boosted-tree families: LightGBM, XGBoost, and CatBoost. We chose this design
because tabular gradient boosting remains highly competitive on heterogeneous
structured features, while being easier to audit than large neural sequence
models.

\subsection{Related Work and SOTA Context}
Our method is closest to the competition-style tabular forecasting tradition:
XGBoost~\cite{chen2016xgboost}, LightGBM~\cite{ke2017lightgbm}, and
CatBoost~\cite{prokhorenkova2018catboost} are strong baselines for mixed
continuous/categorical data and sparse feature interactions. The most relevant
sequence alternatives for this assignment are TCN-style convolutional models
~\cite{bai2018tcn}, PatchTST Transformers~\cite{nie2023patchtst}, and the
simple DLinear baseline~\cite{zeng2023dlinear}, because peer reports explicitly
compared neural sequence models with tree or linear baselines. Recent foundation
models such as TimesFM~\cite{das2024timesfm} show the broader SOTA direction,
but our dataset has a specific supervised target, high-cardinality regions, and
strict leakage constraints, so zero-shot generality is less central than
validation reliability.
In drought forecasting, prior reviews emphasize that meteorological drought is
nonlinear, nonstationary, and often benefits from hybrid or ensemble machine
learning models~\cite{alawsi2022drought}. The U.S. Drought Monitor literature
also motivates our ordinal severity setting: drought categories combine
multiple physical indicators and expert judgment, and transitions are usually
incremental but can change rapidly under heat and rainfall shocks
~\cite{svoboda2002droughtmonitor}.

\section{Proposed Method (\textcolor{red}{1-3 pages})}

\subsection{Problem Formulation}
Let $\mathcal{R}$ be the set of regions and let
$\mathbf{x}_{r,t} \in \mathbb{R}^{D}$ be the meteorological feature vector for
region $r$ on day $t$. Weekly drought labels are observed only on score days,
so most daily rows have missing labels. Given the final test window
$\mathcal{T}_{r}^{test}$ of 91 days for region $r$, we predict five future
weekly scores:
\begin{equation}
  \hat{\mathbf{y}}_{r} =
  (\hat{y}_{r,1},\hat{y}_{r,2},\hat{y}_{r,3},\hat{y}_{r,4},\hat{y}_{r,5}).
\end{equation}
The target metric is
\begin{equation}
  \mathrm{MAE} = \frac{1}{5|\mathcal{R}|}
  \sum_{r\in \mathcal{R}}\sum_{h=1}^{5}
  \left|\hat{y}_{r,h} - y_{r,h}\right|.
\end{equation}

\subsection{Leakage-Aware Feature Construction}
The most important engineering constraint is that score labels inside the
91-day blind window are unavailable at inference time. Therefore all features
that depend on \texttt{score} are computed only from history ending at least
91 days before the forecast origin. This rule also applies to validation:
blind backtests explicitly mask the 91-day window before rebuilding features.
Within this contract, the feature table contains:

\begin{itemize}
  \item \textbf{Weather lags}: shifted daily values at 7, 14, 21, 28, 35, 42,
        and 49 days.
  \item \textbf{Rolling/EWM weather}: rolling mean and standard deviation over
        7--35 days and exponentially weighted means with spans 3, 7, and 14.
  \item \textbf{Long drought proxies}: 30/90/180-day precipitation sums,
        dry-day counts, dry-day share, max dry streak, temperature means,
        dew-point depression, wet-bulb depression, and related heat/dryness
        indices.
  \item \textbf{Calendar and seasonality}: month, quarter, day-of-year,
        week-of-year, and cyclic sine/cosine encodings.
  \item \textbf{Climatology anomalies}: region-month normal weather and
        standardized deviations from those local seasonal norms.
  \item \textbf{Blind-gap score history}: lagged score summaries shifted by at
        least 91 days, so the model sees long-term severity context but not
        hidden test-window labels.
\end{itemize}

Several early experiments used names containing \texttt{two\_stage}; these
names are retained only for artifact compatibility. The active implementation
is a direct-horizon supervised pipeline, not a separate score-reconstruction
system.

\subsection{Model Families}
For each horizon, we train one independent regression model per family rather
than an autoregressive five-week sequence model. For a feature vector
$\mathbf{z}_{r,t}$ at forecast origin $t$, each model family learns an additive
tree function
\begin{equation}
  F_h(\mathbf{z}_{r,t}) =
  F_{0,h}+\sum_{m=1}^{M}\eta_m f_{m,h}(\mathbf{z}_{r,t}),
\end{equation}
where each $f_{m,h}$ is a regression tree fitted to the residual gradient of
the current ensemble. The three libraries share this boosting view, but their
tree structure and regularization are different.

\textbf{LightGBM.} LightGBM is our primary tabular anchor. It first quantizes
continuous inputs into histogram bins, then searches split gains on those bins
instead of on all raw values~\cite{ke2017lightgbm}. Its trees are grown
leaf-wise: after every split, the algorithm expands the current leaf with the
largest estimated loss reduction. This best-first structure can form deep,
specialized rules such as ``very low 90-day precipitation in a specific
seasonal context,'' which is useful for our many lag, rolling, drought-proxy,
and climatology features. The risk is overfitting, so the implementation relies
on constraints such as learning rate, number of leaves, minimum leaf data, and
feature subsampling.

\textbf{XGBoost.} XGBoost also builds an additive ensemble of CART-style
regression trees, but the split search is driven by a second-order regularized
objective~\cite{chen2016xgboost}. At each boosting step it approximates the
loss with gradients and Hessians, penalizes tree complexity through the number
of leaves and leaf weights, and then applies shrinkage and row/column
subsampling. In practice this gives a more conservative, depth-controlled tree
structure than LightGBM. It often captures similar nonlinear interactions, but
with different split choices, making it a strong diversity partner rather than
a duplicate of the LightGBM model.

\textbf{CatBoost.} CatBoost is added because its structure is different in two
ways. First, it uses ordered boosting: training examples are processed through
random permutations so that target-derived statistics for a row are computed
only from earlier rows in that permutation~\cite{prokhorenkova2018catboost}.
This is important because \texttt{region\_id}, month, and other calendar
buckets are high-cardinality categorical signals that can otherwise memorize
labels. Second, CatBoost commonly uses symmetric oblivious trees, where all
nodes at the same depth share the same split. These balanced trees are less
expressive per tree than arbitrary CART trees, but they are stable and provide
a useful categorical-modeling bias for the ensemble.

We used boosted trees deliberately rather than treating them as weak baselines.
The course feedback and peer reports suggest that high-capacity neural models
can look good under an internal metric but fail to transfer when the public and
private distributions differ. Our feature table is heterogeneous, region-heavy,
and partly based on hand-audited historical aggregates. Boosted trees let us
inspect feature ablations, enforce leakage gaps, and keep exact submission
lineage more easily than end-to-end sequence models.

The best reportable submission blends the three model families as
\begin{equation}
  \hat{y}_{r,h}=0.35\hat{y}^{LGB}_{r,h}
  +0.35\hat{y}^{XGB}_{r,h}+0.30\hat{y}^{CAT}_{r,h}.
\end{equation}
All final predictions are clipped to $[0,5]$.

\subsection{Validation and Selection Discipline}
We used four levels of evidence. First, chronological holdout and rolling-origin
validation give fast local feedback. Second, Kaggle-like blind backtests hide a
historical 91-day window and score the next five real weekly labels. Third,
Kaggle public leaderboard readouts check transfer to the competition test
distribution. Fourth, course feedback about public/private gaps was treated as
an overfitting warning: a high public score is useful for placement, but it is
not sufficient evidence that the modeling method generalizes. These signals
often disagreed, so no candidate was promoted from local MAE alone. We also used
a leakage sentinel that poisons hidden-window labels and verifies that
pseudo-test features do not change. This guard was important because a naive
score-persistence model can look excellent locally but fail on Kaggle.

\subsection{Inference and Reproducibility}
At inference time, training tail rows and the 91-day test rows are concatenated
per region, sorted by synthetic date, and transformed using the saved feature
configuration. The last row of the test window becomes the forecast origin for
each region. Every candidate submission is sanity-checked for 2,248 rows, exact
sample-submission column order, matching \texttt{region\_id} order, no NaN
values, and prediction range $[0,5]$. Each important Kaggle score in this
report is tied to a CSV path, Kaggle reference id, blend description, and
lineage label.

\section{Experiments (\textcolor{red}{2-4 pages})}

\begin{table*}[!t]
\centering
\caption{Main Experiment Results and Decisions}
\label{tab:model_results}
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{p{0.15\textwidth}p{0.18\textwidth}p{0.13\textwidth}p{0.08\textwidth}p{0.08\textwidth}p{0.21\textwidth}}
\hline
\textbf{Experiment} & \textbf{Feature / model idea} & \textbf{Validation} &
\textbf{Local MAE} & \textbf{Public MAE} & \textbf{Decision} \\
\hline
Weather-only direct & Direct LGB/XGB without score history & Holdout &
0.6770 / 0.7320 & 0.8640 & Rejected as primary method; weather alone misses
unobserved regional severity context. \\
LGB/XGB 50/50 anchor & Score-history boosted-tree ensemble & Holdout &
0.6942 / 0.7150 & 0.8232 & First reproducible legal anchor. \\
CatBoost tail2737 & Direct CatBoost with native region/calendar categories &
Rolling origin & 0.2212 & Component only & Useful diversity model; validation
scale differs from holdout. \\
\textbf{LGB/XGB/Cat 35/35/30} & Three-model convex ensemble & Kaggle
sanity + lineage audit & N/A & \textbf{0.8124} & \textbf{Best reportable
method claim}, Kaggle ref 52698259. \\
Blind LGBM tail1095 & Refit-full LGBM, lean features & Blind backtest &
0.3549 blind & 0.9380 & Strong pseudo-private anchor but poor public transfer. \\
Feature-fused TCN & 91-day sequence model with visible time-safe priors &
Blind backtest & 0.3508 blind & 0.9450 & Neural diversity signal only; not a
public anchor. \\
Initial accidental hybrid & Leakage-adjacent score behavior & Diagnostic &
0.2100 & 0.8094 & Excluded from reportable claims. \\
Naive score persistence & score gap 0 / forward-fill control & Diagnostic &
0.2321 & 1.0815 & Demonstrates severe leakage risk. \\
\hline
\end{tabular}
\end{table*}

\subsection{Dataset and Setup}
The dataset contains daily meteorological observations over many synthetic
years for 2,248 regions. Training has 5,480 days per region and weekly integer
drought scores; test has 91 unlabeled days per region. Table~\ref{tab:data}
summarizes the data shape.

\begin{table}[h]
\centering
\caption{Dataset Statistics}
\label{tab:data}
\begin{tabular}{lcc}
\hline
\textbf{Property} & \textbf{Train} & \textbf{Test} \\
\hline
Days per region & 5,480 & 91 \\
Regions & 2,248 & 2,248 \\
Rows & 12,319,040 & 204,568 \\
Raw weather columns & 14 & 14 \\
Score range & 0--5 & N/A \\
Score NaN ratio & 85.73\% & N/A \\
\hline
\end{tabular}
\end{table}

Train-test distribution analysis showed that the real test window is not a
simple continuation of the global training distribution. Compared with the
recent 1,095-day training tail, test is hotter by about 5.35 degrees on average
and slightly drier by precipitation and dry-day indicators. Against the stricter
same-region, same-month-day previous-year baseline, daily precipitation drops
from 3.250 to 1.925, dry-day share rises from 0.403 to 0.487, and 90-day
precipitation falls from 277.676 to 223.710. This motivated tail-length,
season-matching, and drought-proxy experiments rather than all-history training.
This directly matches the course suggestion to treat seasonality, recent-year
importance, and data-range choice as first-class modeling decisions. However,
we did not automatically choose the most recent-tail candidate: several
recent-weighted runs improved a local or blind metric but failed to transfer to
the public leaderboard. The final reportable lineage is therefore the candidate
with the best combined evidence, not the candidate with the best single split.

\subsection{Main Model Results}
Table~\ref{tab:model_results} reports the major legal model families and
diagnostic controls. Local MAE values come from different validation protocols,
so they are used only within the same protocol. Kaggle public MAE is the common
external comparison.

The key lesson is that local validation and public transfer are not monotonic.
For example, rolling-origin CatBoost and later neural/TCN variants looked strong
locally, but public leaderboard transfer was much worse than the historical
LGB/XGB/CatBoost anchor. Conversely, the 35/35/30 ensemble made only a modest
modeling change but improved the reproducible public score from 0.8232 to
0.8124.

\subsection{Feature and Validation Ablations}
Feature ablations with LightGBM lean tail1095 clarified which signals were
essential. Removing score-history features degraded rolling MAE by about
0.052, confirming that old severity labels encode persistent regional drought
state and unobserved geography. Removing long drought proxies caused a similar
degradation, showing that accumulated rain deficit, dry streaks, and thermal
stress are not redundant with short lags. Removing climatology improved rolling
MAE by 0.0078, but failed the blind backtest with MAE 0.4463 versus the baseline
blind anchor 0.3549. This was a concrete example where local validation alone
would have selected the wrong model.

The ablations also changed our feature discipline. More features were not
automatically better: target-history features were useful but dangerous, so
they were allowed only behind the 91-day blind gap; climatology features were
kept only when they survived blind checks; and leakage-adjacent artifacts were
excluded even when their public score was attractive. This mirrors the peer
warning that feature quantity can become a liability without validation-based
pruning.

Missingness and shift experiments on 2026-05-23 also failed to replace the
reportable anchor. The best validation/blind candidate,
the season-matched LGBM shift run, reached rolling MAE 0.24574 and blind MAE
0.38370, but the recent pattern was that validation-strong families transferred
poorly to public LB. We therefore kept it diagnostic rather than promoting it
to the main claim.

Table~\ref{tab:param_sensitivity} summarizes the main knobs that we changed
after inspecting either validation failures or live Kaggle feedback. These were
not blind automatic sweeps: each round changed a small number of weights or
feature contracts, then we decided whether the new evidence was reportable,
diagnostic, or only useful for final selection.

\begin{table}[h]
\centering
\caption{Parameter Sensitivity Checks}
\label{tab:param_sensitivity}
\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{tabular}{p{0.26\columnwidth}p{0.34\columnwidth}p{0.30\columnwidth}}
\hline
\textbf{Knob} & \textbf{Values or variants tried} & \textbf{Readout and decision} \\
\hline
CatBoost blend weight & 10, 20, 30, 35, and 40\% CatBoost inside the
LGB/XGB/Cat family & Public MAE was 0.8163, 0.8126, \textbf{0.8124}, 0.8141,
and 0.8168, so 30\% became the reportable anchor. \\
Horizon-specific CatBoost ramp & More CatBoost weight on later forecast weeks &
Public MAE 0.8138; useful robustness probe, but weaker than 35/35/30. \\
Feature groups & Dropped climatology, score history, long drought proxies,
domain indices, and region stats in LGBM lean tail1095 & Score history and long
drought proxies were essential; the climatology drop improved rolling MAE but
failed blind validation. \\
Public-chase horizon alphas & Moved exact public-strong history toward the clean
0.8124 anchor, then adjusted early-week and late-week anchor weights & Public
MAE improved from 0.8049 to 0.7991, then to 0.7930 and finally 0.7905; label as
public-chase, not reportable method evidence. \\
\hline
\end{tabular}
\end{table}

\subsection{Leaderboard and Final-Selection Readout}
Table~\ref{tab:leaderboard} shows the relevant public leaderboard snapshot
from a live Kaggle CLI re-check on 2026-06-04. The public leaderboard is useful
for Kaggle placement, but it can also reward overfitting to public feedback.
Therefore we report two separate artifacts: the reportable model lineage and
the selected public-chase artifact. A 5/29 static private snapshot also showed
Team 5 at rank 5, so the final-selection decision keeps both a public-best
artifact and a near-public private-robust hedge.

\begin{table}[h]
\centering
\caption{Public Leaderboard Snapshot}
\label{tab:leaderboard}
\footnotesize
\begin{tabular}{lcc}
\hline
\textbf{Entry} & \textbf{Public MAE} & \textbf{Rank} \\
\hline
Team 10 & 0.7371 & 1 \\
Team 2 & 0.7598 & 2 \\
Team 3 & 0.7628 & 3 \\
Team 15 & 0.7754 & 4 \\
Team 4 & 0.7810 & 5 \\
Team 20 & 0.7836 & 6 \\
Team 14 & 0.7904 & 7 \\
\textbf{Team 5 public-chase} & \textbf{0.7905} & \textbf{8} \\
Team 1 & 0.7909 & 9 \\
Team 11 & 0.8010 & 10 \\
Team 8 & 0.8027 & 11 \\
Baseline 3 & 0.8056 & 12 \\
Team 5 reportable lineage & 0.8124 & -- \\
Baseline 2 & 0.8623 & -- \\
Baseline 1 & 0.9117 & -- \\
\hline
\end{tabular}
\end{table}

The selected public-chase artifact is
\path{submissions/baseline3_private_hedge_v8_cat35_w1_0p0375_keep_shape_horizon_0p0375_0p375_0p55_0p75_1.csv}
(Kaggle ref 53204258, SHA-12 390565517cb3, public MAE 0.7905). It is an
exact-history horizon hedge moving the recovered 0.8094 public reference toward
the clean 0.8124 anchor by horizon alphas
$[0.0375,0.375,0.55,0.75,1.00]$. It is useful for final selection, but it is
not a new reportable training algorithm. The current near-public private-risk
alternative selected for the second Static Private slot is
\path{submissions/baseline3_private_hedge_v9_cat35_w1_0p075_w2_0p425_stronger_late_anchor_horizon_0p075_0p425_0p6_0p825_1.csv}
(Kaggle ref 53259683, SHA-12 262f683b87b6, public MAE 0.7909). Its horizon
alphas are $[0.075,0.425,0.60,0.825,1.00]$, so it gives up only 0.0004
displayed public MAE while moving week 2 and later horizons closer to the clean
anchor. The v8 private hedge ref 53204319 remains the pure-delta fallback, and
the previous v7 pair, refs 53186508 and 53186571, remains a fallback if the
team wants the earlier manual selection.

Because private leaderboard risk remains possible, the post-crossing workflow did
not stop at the public crossing. The v9 follow-up consumed the expanded
10/day quota on 2026-06-01 after all files passed sanity checks; the current
manual final-selection recommendation is to select refs 53204258 and 53259683.
Ref 53038036 remains a stronger private fallback if a larger public gap is
acceptable. These files are sanity-checked final-selection artifacts only, not new
reportable model claims.

The reportable artifact remains
\path{submissions/ensemble_20260516_lgb_xgb_cat2737_35_35_30.csv}
(Kaggle ref 52698259, SHA-12 bee6f618828d when restored locally). Its
prediction distribution is valid and conservative: week means are approximately
0.955, 0.927, 0.880, 0.831, and 0.817, and all predictions are clipped within
$[0,5]$.

\subsection{Discussion}
The experiments reveal a ``Kaggle paradox.'' Pure weather models and some
validation-strong models look reasonable locally but underperform on the public
leaderboard. Historical score features, when shifted safely beyond the 91-day
blind gap, provide information about persistent regional vulnerability that raw
weather alone cannot reconstruct. However, target history also creates the
largest leakage risk, so the final workflow treats leakage sentinels, blind
backtests, and exact artifact lineage as part of model selection rather than as
after-the-fact documentation.

Peer reports support the same conclusion from different angles. One group found
that discarding older years improved public score but raised public/private
overfitting concern; another used LightGBM, TCN, and feature-fused TCN as an
ensemble, where LightGBM still added marginal value despite not being the main
sequence model; a third group tried GRU, LSTM, CNN, Transformer, and tree
models, but reported that neural models did not transfer as reliably as their
internal metrics suggested. Our response is therefore not to maximize model
complexity, but to emphasize distribution checks, leakage-safe features,
heterogeneous tree ensembles, and conservative artifact labeling.

The public-chase sequence further shows why leaderboard scores must be labeled
carefully. It improved Team 5 from the reportable 0.8124 anchor to 0.7905 by
using public feedback around exact historical artifacts. This is acceptable for
Kaggle final selection if the competition UI allows selecting the submitted
artifact, but it should not be described as evidence that the supervised
training method itself reached 0.7905.

In plain language, our final interpretation is that the strongest transferable
modeling idea is still the leakage-safe boosted-tree ensemble, while the best
Kaggle score comes from small, manually reviewed horizon-weight adjustments
around known submitted shapes. That is why the report keeps the 0.7905 and
0.8124 numbers side by side instead of merging them into one claim.

\section{Conclusion}
We built a leakage-aware drought severity forecasting pipeline for the Data
Mining final project. The final reportable method uses direct-horizon
LightGBM, XGBoost, and CatBoost models over carefully shifted temporal,
meteorological, climatology, and historical severity features. The best
reproducible modeling claim is the 35/35/30 boosted-tree ensemble with public
MAE 0.8124. Extensive diagnostics showed that score history and long drought
proxies are essential, that local rolling-origin MAE can be misleading under
train-test shift, and that blind-window leakage controls are necessary. A
separate public-chase artifact reaches public MAE 0.7905 and crosses Baseline
3; we keep it available for final leaderboard selection while preserving the
clean 0.8124 lineage as the scientific method claim.

\bibliographystyle{IEEEtran}
\bibliography{reference}

\end{document}
